PyTracer: Automatically profiling numerical instabilities in Python

### 2021-10-28 paper.tex
First complete draft

\documentclass[11pt]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{todonotes}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{makecell}
\usepackage{ulem} % provides sout
\usepackage{minted} % Provides listing environment

@@

\begin{abstract}
[abstract goes here]
\end{abstract}

@@

The scientific Python ecosystem is a central component of scientific
analyses due to its rich data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Diverse research groups built numerous domain-specific data analysis pipelines from this ecosystem, leveraging Python's simplicity and expressivity.

@@

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces, enabling scaling to large programs without manual intervention. In contrast, theoretical approaches based on condition number or backward error analyses require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques such as Frama-C~\cite{cuoq2012frama}, Gappa~\cite{de2010certifying} or Flocq~\cite{boldo2011flocq} hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

PyTracer estimates the significant digits of Python floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) tracing floating-point computations along program executions, (2) generating numerically-perturbed executions, and (3) visualizing stability evaluations. PyTracer addresses these requirements with:
Dynamic instrumentation of Python functions, modules, and classes through meta-programming
A ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic
An interactive Plotly dashboard to visualize stability measures throughout program executions

Through stability evaluations of 
SciPy, scikit-learn, and pyAFQ~\cite{kruper2021evaluating} --- an analysis toolbox for diffusion magnetic resonance imaging --- we demonstrate PyTracer as a practical solution to review the numerical stability of extensive code bases. The remainder of this manuscript describes related approaches, presents the design of PyTracer, and demonstrates its potential in various use cases.

\section{Related work}

Several tools have been and continue to be developed to assess numerical quality. Those tools can be divided into two main categories depending on whether they adopt a static or a dynamic approach.
The static analysis produces theoretically valid error bounds, while dynamic analysis is generally more scalable, although it only provides experimental estimations.
\pytracer adopts the dynamic approach due to its scalability to large existing code bases. Dynamic approaches typically rely on the tracing of floating-point computations to detect instabilities and visualize them. The remainder of this section reviews existing approaches for each of these three steps.

@@

\label{sec:soa}
\subsection{Numerical tracing techniques}
Tools to trace floating-point computations for C, C++, or FORTRAN programs are plethora due to the prevalence of these languages in High-Performance Computing (HPC). 
The main tracing techniques are source-to-source, compiler-based transformations, and dynamic binary instrumentation.

The Source-to-source technique requires a rewriting of the application to use modify floating-point types.
CADNA~\cite{jezequel2008cadna} is a library for C, C++ and Fortran  implementing the CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
pLiner uses source-to-source transformation at the Abstract Syntax Tree (AST)  level to rewrite parts of code with higher precision. 
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
MCAlib~\cite{frechtling2015mcalib} is a C library that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using the MPFR~\cite{fousse2007mpfr} library.
The work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code in infinite and fixed precision with and without using stochastic arithmetic.

The compiler-based approach uses a compiler to analyze or replace floating-point expressions with other models automatically. 
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including MCA. 
The work in ~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root cause analyzer based on the Clang compiler that detects floating-point operations responsible for result variability. 
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel}, is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability induced by compilers and their optimizations.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB, respectively.

Dynamic Binary Instrumentation (DBI) operates on an executable directly, without the need for recompilation or manual changes, and is thus
transparent to the programming language used. However, working at a binary level makes it difficult to access high-level information needed
to debug and understand the source code logic. In particular, analyzing Python code with DBI is feasible but provides minimal insights on the stability of specific Python code sections.
CraftHPC~\cite{lam2013dynamic} uses DBI to detect catastrophic cancellations at runtime.
Verrou~\cite{fevotte2016verrou} is a Valgrind~\cite{nethercote2007valgrind} based tool that replaces on-the-fly
floating-point operations by their stochastic arithmetic counterparts. FPDebug~\cite{benz2012dynamic} uses DBI to detect numerical inaccuracies by using shadow computations with higher precision.
Herbgring~\cite{sanchez2017finding} is a Valgrind-based tool to detect
numerical instabilities. It uses a shadow memory to detect precision losses by comparison with results obtained with the high-precision 
MPFR library~\cite{fousse2007mpfr}. It is combined with symbolic computation to backtrack the root of the error.
We can note that all methods based on a high-precision oracle suppose that computing with a large number of bits is sufficient to obtain an accurate result, which is not always true, see, for example, the famous Muller's sequence ~\cite{chatelain2018veritracer}.

Being agnostic to the programming language is a serious advantage of the DBI method compare to source-to-source and compiler-bases methods. However, working at a binary level makes it difficult to access high-level information needed to debug and understand the source code logic. On the other hand, Source-to-source presents the advantage of having a fine-grained control on code analyzed but lacks scalability, as rewriting large codes can be a tedious task if done by hand. Finally, the compiler-based approach takes advantage of the best of both by being automatic and having access to high-level information. However, like source-to-source, the compiler-based method is not suitable for analyzing closed-source proprietary libraries.

To the best of our knowledge, there is currently no tool dedicated to the numerical analysis of Python code. Existing tools for tracing Python code are dedicated to performance profiling, for time (cProfile) 
or memory consumption (memprofile). 
Anteater~\cite{faust2019anteater} is a Python tracer tool that aims at debugging Python applications. 
It operates source transformation at the AST level but only deals with native numeric Python types.
Moreover, the user needs to tag each variable to trace manually. Finally, according to the authors, Anteater does not scale to large traces.
However, we can compare the methodology used by cProfile, memprofile, and Anteater against the one used in Pytracer.
These tools use instrumentation based on decorators, a Python mechanism allowing to instrument a function by adding a line over a function declaration.
While this method is proper when targeting specific functions, it becomes time-consuming when applied to many functions since one needs to tag functions manually, which is not feasible for large codebases where potentially unstable code sections are unknown.

@@

Stochastic arithmetic is a theoretical model that exploits randomness to estimate the floating-point format's numerical instabilities. The idea is to treat round-off errors as a random variable for then using classical statistical tools to study them.

The first published complete stochastic arithmetic framework is CESTAC~\cite{vignes1993stochastic}. Each floating-point computation is run multiple times in this method, switching the rounding mode to either round-up or round-down with
an equal probability. From these stochastic samples, one can then derive the number of significant digits in the computed result. In addition, the Discrete Stochastic Arithmetic (DSA~\cite{vignes2004discrete}) extends the CESTAC method to comparison operators implemented in the CADNA library.

The Monte Carlo Arithmetic works on the same principle but introduces two differences:
a virtual precision parameter allowing to simulate reduced working precision and different \texttt{modes} to introduce perturbations
on the inputs or the output.
MCA simulates round-off errors by introducing random perturbations in floating-point computations and replaces each floating-point number with its stochastic counterpart:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the basis, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ a random variable.
Virtual precision is one strength of the MCA method since it allows simulating reduced working precision.
MCA comes with three modes: Random Rounding (RR), Precision Bounding (PB), and FullMCA, which perturb either the output, the inputs, or both with the $inexact$ function. While the RR mode is equivalent to stochastic rounding, PB mode allows identifying catastrophic cancellations.

MCA provides a formula \tristan{is this formula specific to MCA or can it also be applied to CESTAC?} \Yohan{Yes, CESTAC supposes that the random variable is normal and so use a different formulation} to compute the number of significant digits:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\sigma$ and $\mu$ are respectively the empirical standard deviation and expected value of a variable $x$ evaluating 
several times with MCA. 
Note that this formula gives an approximation of the real. Recently, Sohier et al. ~\cite{sohier2018confidence} introduced a 
general theoretical framework that includes DSA and MCA formulae and allows to obtain associated rigorous confidence intervals on the number of significant digits estimated.

\subsubsection{Uncertainty and sensitivity analysis}

Uncertainty analysis (UA) is a method to derive the output distribution response from an input distribution.
The idea is to model uncertainty on data due to limited precision measurement, for example, by a probability distribution, and to estimate the impact on the result. 
Sensitivity analysis (SA) is quite close to uncertainty analysis but differs from the scale observed.
While uncertainty analysis attempts to describe the whole output distribution, sensitivity analysis measures the local response to a small perturbation~\cite{loucks2017water}. 
Several approaches for UA and SA exist, among them differential analysis, response surface methodology, Monte Carlo analysis, and variance decomposition (see~\cite{helton2006survey}). In particular, Monte Carlo Analysis is a sampling-based technique that samples inputs space to estimate output response. Monte Carlo analysis shares similarities with Monte Carlo Arithmetic. For instance, the PRECISE framework~\cite{chaitin1996lectures} shares with MCA the idea of varying virtual precision to estimate output sensitivity.

\subsubsection{Random seed sensitivity}

The quality of a Random Number Generator is a crucial characteristic when dealing with stochastic model and differ from cryptography requirements. For numerical simulation, an RNG of quality must have a good tradeoff between performance, period, correlation, or uniformity~\cite{james2020review}. Poor quality RNGs may have a substantial impact on results~\cite{click2011quality}. First, measuring variation induced by seed is a good indicator of program stability that we can compare against variability induced by floating-point arithmetic.
\Yohan{This section lacks solid arguments. Needs improvement}

@@

The visualization covers a wide range of specific domains in the numerical simulation area, from general-purpose numerical visualization with tools
like VTK~\cite{schroeder2000visualizing}, to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
We review here tools dedicated to the analysis of the numerical variability induced by the floating-point model. We can first highlight that despite the large number of existing tools dedicated to pinpoint source code instructions, few of them allow the user to have a global view of the numerical accuracy of their computations or compare several executions of the same program under various perturbations.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer to visualize the numerical stability over time for C, C++, and Fortran codes. It is a trace-based tool built as an extension of Verificarlo~\cite{verificarlo} that offers contextual information about variables traced to allow the user to identify instabilities quickly.
However, it does not offer a dynamic interface neither takes into account information about high-level structures like \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} is a tool for visualizing the effects of floating-point arithmetic but only for educational purposes.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python codes, but as mentioned previously, it only deals with native Python types, and the user needs to select variables to trace manually. However, Anteater provides an accurate reflection around visual debugging beyond floating-point analysis, which has inspired the design of \pytracer's visualization interface.

@@

The PyTracer workflow follows a three-step approach (Figure~\ref{fig:workflow}). First, when a Python application is loaded, \pytracer instruments using ``monkey patching", a technique to modify Python programs without editing their source code. The instrumentation replaces all the functions with a wrapper that saves their arguments and outputs in a trace file using ``pickle", Python's object serialization mechanism. \pytracer then aggregates trace files obtained from multiple perturbed executions, computing summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow \tristan{move the execution and noise injection to step 2}}
    \label{fig:workflow}
\end{figure}

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the original application can transparently use it.
By default, \pytracer instruments all the attributes
in a given module \tristan{not only the functions?} \Yohan{No, you can have class definitions or instances that need to be wrapped}, except the special ones of the form \texttt{\_\_*\_\_}.
However, some attributes are not writable; in this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of attributes to trace through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches, which enable instrumentation in read-only environments such as Singularity containers or HPC clusters.

@@

Python implements default \texttt{finder} and \texttt{loader} classes and search paths for importing a module, but it also provides a way to create custom \texttt{finder} and \texttt{loader} classes.
To register a new \texttt{finder} class, one needs to add it to the \texttt{sys.meta\_path} list.
When Python encounters an \texttt{import} statement, it first looks for a binding in the \texttt{sys.modules}
and then iterates over \texttt{sys.meta\_path}.
Python queries each \texttt{finder} class successively until it finds the module or reaches the end of the list, raising a \texttt{ModuleNotFoundError}. Then, \pytracer adds a custom \texttt{finder} class at the head of the list that intercepts
the module import and creates the module with a custom \texttt{loader} class.

\pytracer's \texttt{loader} class first loads the original module then copies it as a new instance of the \texttt{ModuleType} class. It then calls the appropriate instrumentation function for each module attribute depending on its type (function, class, or instances). Next, \pytracer replaces the original module with its instrumented counterpart in the \texttt{sys.modules} map once it has instrument all the attributes. Hence, the calling convention remains unchanged, and the application will transparently call the instrumented module. Finally, Pytracer updates all references to the original module that an object may hold. For example, functions have a \texttt{\_\_globals\_\_} attribute that contains references to internal objects that the function needs. Since those references point to non-instrumented objects, Pytracer must update them.
\tristan{is the non-instrumented version ever added to sys.modules? If so, could this lead to race conditions? If not, replace with ``adds the instrumented module to the sys.modules map"}. 
\Yohan{Yes, it replaces the original module. You can't have a race-condition since sys.modules is a dict with only one value for a given key. The trick is that the wrapped module keeps a reference to the original module, it's just that this reference is invisible for the others object since it's not in sys.modules and globals. By the way, Pytracer needs to update all possible references that original objects may hold, like functions in \_\_dict\_\_ attributes or locals references in \_\_globals\_\_ for example.}

@@

Module functions are the most straightforward case to treat.
For each function, \pytracer's loader class creates a wrapper function, dynamically compiles it with the compile builtin function, and substitutes it for the original module function.
However, we cannot apply this technique to callable class instances, i.e., class instances that have the \texttt{\_\_call\_\_} attribute but are not of the \texttt{FunctionType} type. Indeed, applying the wrapper technique would 1) modify the type of the class instance to \texttt{FunctionType}, which could cause syntactic and semantic bugs, and 2) mask class attributes, leading to \texttt{AttributeError} exceptions.
To overcome this issue, \pytracer overrides the \texttt{\_\_call\_\_} attribute with the wrapper function when possible. When the \texttt{\_\_call\_\_} attribute is not writable,  Pytracer does not instrument the class and returns a warning. 
Listings~\ref{fig:wrapper_creation} and ~\ref{fig:generic_wrapper} show \pytracer's wrapper creation function.

\tristan{add docstrings to all Python code examples and clarify captions accordingly: you don't need to explain the meaning of each attribute in the caption if it's in the docstring. The caption can focus on the main functionality, reusing the text currently in 3.1.5.}
\Yohan{TODO}

Listing~\ref{fig:wrapper_creation} shows how functions are instrumented.
This function takes as inputs the information about the function to instrument (module name and full qualified name),
the function itself, and the actual name of the function. It then creates a wrapper around the actual wrapper
(\texttt{generic\_wrapper}) that will dump the values. Instead of passing the function, we pass the function's identifier, which is a unique 64-bits integer that references each Python object. This identifier is available
with the \texttt{id} builtin function. it allows to get rid of aliases and to avoid instrumenting functions twice.

Listing~\ref{fig:generic_wrapper} shows what the instrumented code does. 
It first unpacks the function's information to get the function's identifier to retrieve the original function.
Then arguments are bound by creating a mapping from positional and keyword arguments to parameters, avoiding conflicts and mispositioning. It also allows getting names of arguments for the visualization.
Once bound, the \texttt{inputs} function dumps the arguments in the pickle file.
Finally, Pytracer calls the function with the correct arguments, dumps the result, and returns it.

@@

\subsection{Overhead}

>>>>

\documentclass[11pt]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{todonotes}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{makecell}
\usepackage{ulem} % provides sout
\usepackage{minted} % Provides listing environment
\usepackage[htt]{hyphenat}
\usepackage{subcaption}
% \usepackage{subfigure}

@@

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computational codes. Despite the recent large numbers of tools dedicated to evaluating codes' numerical stability, analyzing a complete program remains a handy task for the user. To address this issue, we propose \pytracer, an automatic profiler for analyzing numerical instabilities of Python applications. \pytracer automatically instruments Python code to produce numerical traces that can be visualized through a Dash server, allowing an interactive visualization of numerical stability.
\pytracer is designed to work with any random noise model that simulates numerical instabilities, and we demonstrate its usage with the Monte Carlo Arithmetic through the fuzzy ecosystem.
Finally, we demonstrate that pytracer can handle intensively used Python scientific libraries such as NumPy, SciPy, and Scikit-learn on a representative set of tests, at low cost with an average slowdown $\times 3.5$.
\end{abstract}

@@

The scientific Python ecosystem is a central component of scientific
analyses due to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Diverse research groups built numerous domain-specific data analysis pipelines from this ecosystem, leveraging Python's simplicity and expressivity.

@@

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces, enabling scaling to large programs without manual intervention. In contrast, theoretical approaches based on condition numbers or backward error analysis require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques such as Frama-C~\cite{cuoq2012frama}, Gappa~\cite{de2010certifying} or Flocq~\cite{boldo2011flocq} hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

PyTracer estimates the number of significant digits of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) tracing floating-point computations along program executions, (2) generating numerically-perturbed executions, and (3) visualizing stability evaluations. PyTracer addresses these requirements with (1) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (2) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

Through stability evaluations of 
SciPy, scikit-learn, and PyAFQ~\cite{kruper2021evaluating} --- an analysis toolbox for diffusion magnetic resonance imaging --- we demonstrate PyTracer as a practical solution to review the numerical stability of extensive code bases. The remainder of this manuscript describes related approaches, presents the design of PyTracer, and demonstrates its potential in various use cases.

\section{Background and related work}

Several tools have been and continue to be developed to assess numerical quality. Those tools can be divided into two main categories depending on whether they adopt a static or a dynamic approach.
Static analysis produces theoretically valid error bounds, while dynamic analysis is generally more scalable, although it only provides experimental estimations.
\pytracer adopts the dynamic approach due to its scalability to large existing code bases. Dynamic approaches typically trace floating-point computations, detect instabilities in traces, and visualize summary statistics about instabilities. The remainder of this section reviews existing approaches for each of these steps.

@@

\label{sec:soa}
\subsection{Numerical tracing techniques}
Tools to trace floating-point computations for C, C++, or FORTRAN programs are numerous due to the prevalence of these languages in High-Performance Computing (HPC). 
The main tracing techniques are source-to-source, compiler-based transformations, and dynamic binary instrumentation.

The source-to-source technique requires a rewriting of the application to modify floating-point types. Among frameworks that adopted this technique,
CADNA~\cite{jezequel2008cadna} is a library for C, C++ and Fortran implementing the CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
MCAlib~\cite{frechtling2015mcalib} is a C library that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using the MPFR~\cite{fousse2007mpfr} library.
Finally, the work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code in infinite and fixed precision with and without  stochastic arithmetic.

The compiler-based approach uses a compiler to analyze or replace floating-point expressions with other models automatically. 
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including MCA. 
The work in~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root cause analyzer based on the Clang compiler that detects floating-point operations responsible for result variability using a source-to-source transformation at the Abstract Syntax Tree (AST)  level to rewrite parts of code with higher precision. 
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel}, is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability induced by compilers and their optimizations.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB, respectively.

Dynamic Binary Instrumentation (DBI) operates on an executable directly, without the need for recompilation or manual changes, and is thus
transparent to the programming language used. 
CraftHPC~\cite{lam2013dynamic} uses DBI to detect catastrophic cancellations at runtime.
Verrou~\cite{fevotte2016verrou} is a Valgrind~\cite{nethercote2007valgrind} based tool that replaces on-the-fly
floating-point operations by their stochastic arithmetic counterparts. FPDebug~\cite{benz2012dynamic} uses DBI to detect numerical inaccuracies by using shadow computations with higher precision.
Herbgring~\cite{sanchez2017finding} is a Valgrind-based tool to detect
numerical instabilities. It uses a shadow memory to detect precision losses by comparison with results obtained with the high-precision 
MPFR library~\cite{fousse2007mpfr}. It is combined with symbolic computation to backtrack the root of the error.
We can note that all methods based on a high-precision oracle suppose that computing with a large number of bits is sufficient to obtain an accurate result, which is not always true, see, for example, the famous Muller's sequence ~\cite{chatelain2018veritracer}.

Being agnostic to the programming language is a serious advantage of the DBI method compared to source-to-source and compiler-bases methods. However, working at a binary level makes it difficult to access high-level information needed to debug and understand the source code logic. Conversely, the source-to-source approach provides a fine-grained control on the code being analyzed, bu itt lacks scalability, as rewriting large codes can be a tedious task. Finally, the compiler-based approach takes advantage of the best of both by being automatic and having access to high-level information. However, like source-to-source, the compiler-based method is not suitable for analyzing closed-source libraries.

To the best of our knowledge, there is currently no tool dedicated to the numerical analysis of Python code. Existing tools for tracing Python code are dedicated to performance profiling, for time (cProfile) 
or memory consumption (memprofile). 
Anteater~\cite{faust2019anteater} is a Python tracer tool to debug Python applications. 
It performs source transformations at the AST level but only deals with native numeric Python types.
Moreover, the user needs to manually tag each variable to trace. Finally, according to the authors, Anteater does not scale to large traces.
cProfile, memprofile, and Anteater use Python decorators as the main instrumentation technique, a Python mechanism to instrument a function by adding a line over a function declaration.
While this method is appropriate when targeting specific functions, it is not feasible for large code bases where the location of potentially unstable code sections is unknown.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities coming from the use of floating-point representations. The main idea is to treat round-off errors 
% \tristan{only round-off errors or all numerical errors?} 
as random variables and to characterize them statistically.

The first published complete stochastic arithmetic framework was CESTAC~\cite{vignes1993stochastic}, where each floating-point computation is performed multiple times with different rounding modes: round-up or round-down. From these stochastic samples, one can then derive the number of significant digits in the computed result. Discrete Stochastic Arithmetic~\cite{vignes2004discrete} extends CESTAC to comparison operators implemented in the CADNA library.

Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} uses the same principle but introduces two differences:
(i) a virtual precision parameter, allowing to simulate reduced working precision,  and (ii) different perturbation modes, to introduce perturbations
on function inputs or output.
MCA simulates round-off errors by introducing random perturbations in floating-point computations. It replaces each floating-point number with its stochastic counterpart:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a uniform variable.
Virtual precision allows for the simulation of reduced working precision.
MCA can be applied in three modes: Random Rounding (RR), Precision Bounding (PB), and full MCA, which respectively perturb either the output, the inputs, or both with the $inexact$ function. While the RR mode is equivalent to stochastic rounding, the PB mode allows identifying catastrophic cancellations.

In stochastic arithmetic, numerical error is measured as the number of significant digits $s$ estimated among the sampled values. A common formula to determine this number from MCA samples is the one reported in~\cite{parker1997monte}:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a variable sampled with MCA.  Recent variations of this formula that also provide confidence intervals are discussed in~\cite{sohier2018confidence}.

\subsubsection{Uncertainty, sensitivity, and random seed analysis}

Uncertainty analysis (UA) is a method to estimate the output distribution given an input distribution.
Sensitivity analysis (SA) is a variation of UA that measures the local response to a small perturbation~\cite{loucks2017water}. 
Several approaches for UA and SA exist, among them differential analysis, response surface methodology, Monte Carlo analysis, and variance decomposition (see~\cite{helton2006survey} for a review). In particular, Monte Carlo analysis is a sampling-based technique that samples inputs space to estimate output response. Monte Carlo analysis shares similarities with Monte Carlo Arithmetic. For instance, the PRECISE framework~\cite{chaitin1996lectures} shares with MCA the idea of varying virtual precision to estimate output sensitivity.
Furthermore, the quality of a Random Number Generator is a crucial characteristic when dealing with stochastic model and differ from cryptography requirements. For numerical simulation, a good RNG provides a tradeoff between performance, period, correlation, or uniformity~\cite{james2020review}. On the other hand, poor quality RNGs may have a substantial impact on results~\cite{click2011quality}. First, measuring variation induced by seed is a good indicator of program stability that we can compare against variability induced by floating-point arithmetic.

@@

Visualization covers a wide range of numerical simulation domains, from general-purpose numerical visualization with tools
such as VTK~\cite{schroeder2000visualizing}, to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
We review here tools dedicated to the analysis of numerical variability induced by the floating-point model. Despite the large number of existing tools  to locate unstable source code sections, few of them allow the user to have a global view of the numerical accuracy of their computations or compare several executions of the same program under various perturbations.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer of numerical stability over time for C, C++, and Fortran codes. It is an extension of Verificarlo~\cite{verificarlo} that offers contextual information about variables traced to allow the user to identify instabilities quickly.
However, it does not offer a dynamic interface and does not include information about high-level structures such as \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} visualizes the effects of floating-point arithmetic but only for educational purposes.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python codes, but as mentioned previously, it only deals with native Python types, and the user needs to select variables to trace manually. However, Anteater provides an insightful reflection 
% \tristan{It's a bit vague. Do you simply mean: ``Anteater's GUI inspired the design of \pytracer's interface"?} 
around visual debugging beyond floating-point analysis, which has inspired the design of \pytracer's visualization interface.

@@

The PyTracer workflow follows a three-step approach (Figure~\ref{fig:workflow}). First, \pytracer instruments applications using ``monkey patching", a technique to modify Python programs without editing their source code. The instrumentation replaces all the functions with a wrapper that saves their arguments and returned value in a trace file using ``pickle", Python's object serialization mechanism. \pytracer then aggregates trace files obtained from multiple perturbed executions, computing summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow}
    \label{fig:workflow}
\end{figure}

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the original application can transparently use it.
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}.
However, some attributes are not writable; in this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches, which is scalable and useful in read-only environments such as Singularity containers or HPC clusters.

@@

Python supports custom \texttt{finder} classes, registered in the \texttt{sys.meta\_path} list.
When Python encounters an \texttt{import} statement, it first looks for a binding in the \texttt{sys.modules}
and then iterates over the \texttt{finder} classes in \texttt{sys.meta\_path} until it finds the module to import. \pytracer adds a custom \texttt{finder} class at the head of the list that intercepts
the module import and creates the module with a custom \texttt{loader} class.

\pytracer's \texttt{loader} class first loads the original module, then copies it as a new instance of the \texttt{ModuleType} class. It then calls the appropriate instrumentation function for each module attribute depending on its type (function, class, or instances). Then, \pytracer replaces the original module with its instrumented counterpart in the \texttt{sys.modules} map once it has instrument all the attributes. Hence, the calling convention remains unchanged, and the application will transparently call the instrumented module. Finally, Pytracer updates all existing references to the original module, such as the ones in \texttt{\_\_globals\_\_} function attributes that contain references to internal function objects.

% \tristan{is the non-instrumented version ever added to sys.modules? If so, could this lead to race conditions? If not, replace with ``adds the instrumented module to the sys.modules map"}. 
% \Yohan{Yes, it replaces the original module. You can't have a race-condition since sys.modules is a dict with only one value for a given key. The trick is that the wrapped module keeps a reference to the original module, it's just that this reference is invisible for the others object since it's not in sys.modules and globals. By the way, Pytracer needs to update all possible references that original objects may hold, like functions in \_\_dict\_\_ attributes or locals references in \_\_globals\_\_ for example.}

@@

For each function, \pytracer's \texttt{loader} class creates a wrapper function, dynamically compiles it with the compile builtin function, and substitutes it for the original module function. This simple technique
cannot be applied to callable class instances, i.e., class instances that have the \texttt{\_\_call\_\_} attribute but are not of the \texttt{FunctionType} type. Indeed, applying the wrapper technique would 1) modify the type of the class instance to \texttt{FunctionType}, which could cause syntactic and semantic bugs, and 2) mask class attributes, leading to \texttt{AttributeError} exceptions.
To overcome this issue, \pytracer overrides the \texttt{\_\_call\_\_} attribute with the wrapper function when possible. When the \texttt{\_\_call\_\_} attribute is not writable,  Pytracer does not instrument the class and returns a warning. 
Listings~\ref{fig:wrapper_creation} and~\ref{fig:generic_wrapper} show \pytracer's wrapper creation function.

% \tristan{add docstrings to all Python code examples and clarify captions accordingly: you don't need to explain the meaning of each attribute in the caption if it's in the docstring. The caption can focus on the main functionality, reusing the text currently in 3.1.5.}
% \Yohan{TODO}

% Listing~\ref{fig:wrapper_creation} shows how functions are instrumented.
% This function takes as inputs the information about the function to instrument (module name and full qualified name),
% the function itself, and the actual name of the function. It then creates a wrapper around the actual wrapper
% (\texttt{generic\_wrapper}) that will save the values in trace files. Instead of passing the function, we pass the function's identifier, which is a unique 64-bits integer that references each Python object. This identifier is available
% with the \texttt{id} builtin function. it allows to get rid of aliases and to avoid instrumenting functions twice.  \tristan{Summarize this paragraph in the figure caption}

% Listing~\ref{fig:generic_wrapper} shows what the instrumented code does. 
% It first unpacks the function's information to get the function's identifier to retrieve the original function.
% Then arguments are bound by creating a mapping from positional and keyword arguments to parameters, avoiding conflicts and mispositioning. It also allows getting names of arguments for the visualization.
% Once bound, the \texttt{inputs} function dumps the arguments in the pickle file.
% Finally, Pytracer calls the function with the correct arguments, dumps the result, and returns it. \tristan{Summarize this paragraph in the figure caption}

@@

\subsection{Tracing}
